SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2018-09-28 scipy-1.0/paper.tex
Revise background

With a rich programming environment and a numerical array object in
place, the time was ripe for the development of a full scientific
software stack. Many of the first efforts were driven by graduate
students and post-doctoral researchers, but soon attracted the
attention of more senior researchers.
% Paul Dubois, CDAT
% https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html
For example, Paul Dubois at Lawrence Livermore National Laboratory took over the
maintenance of Numeric and funded the writing of its
manual\footnote{http://www.numpy.org/\_downloads/numeric-manual.pdf}, and
%in the Program for Climate Model Diagnosis and Intercomparison at
%Lawrence Livermore National Laboratory built their Climate Data
%Analysis Tools (``an open-source set of Python modules and tools for
%accessing, manipulating, and plotting scientific data in general and
%climate data sets in
%particular''\footnote{https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html})
%on Numeric, the newly established Python array object.  Paul later
%became the maintainer for Numeric, and helped fund the writing of the
%first manual\footnote{https://conference.scipy.org/scipy2010/slides/travis_oliphant_keynote.pdf}
in 1998 the Space Telescope Science Institute, in charge of science
operations of the Hubble Space Telescope, decided to replace their
custom scripting language with Python\footnote{https://conference.scipy.org/scipy2011/slides/greenfield\_keynote\_astronomy.pdf}.

% Perry Greenfield, Hubble
% https://conference.scipy.org/scipy2011/slides/greenfield_keynote_astronomy.pdf
%
% Note: plotting was important for them, so they eventually
% funded development on Matplotlib.

@@

In 2001, Eric Jones and Travis Vaught founded Enthought Scientific
Computing Solutions (now Enthought, Inc.) in Austin, Texas.  In an
effort to simplify the tool stack, they created the SciPy project,
centered around the SciPy library which would subsume all the
above-mentioned packages.
The new project quickly gained momentum, with a website and code
repository appearing in
February\footnote{https://web.archive.org/web/20010309040805/http://scipy.org:80/}, and
a mailing list announced in
June \footnote{https://mail.python.org/pipermail/scipy-dev/2001-June/000000.html}. By
August, a first release was announced\footnote{https://mail.python.org/pipermail/python-list/2001-August/106419.html}, an excerpt of which is shown in
Fig.~\ref{fig:announce-0.1}.
In September, the first documentation was
published\footnote{https://web.archive.org/web/20021013204556/http://www.scipy.org:80/scipy/site\_content/site\_news/docs\_released1},
and by the following September the inaugural SciPy workshop was held
at Caltech\footnote{https://mail.python.org/pipermail/numpy-discussion/2002-June/001511.html}.

@@

At this point, scientific Python is starting to attract more serious
attention; code that started out as side projects by graduate students
has grown into essential infrastructure at national laboratories and
research institutes.  But even as SciPy, the algorithms library of the
ecosystem, started to mature, the next generation of graduate students
and postdocs were already exploring other aspects of the computational
environment.

In 2000, Prabhu Ramachandran\fixme{also note that he, and the others
  mentioned below, are also SciPy contributors}, a PhD student at the Indian Institute of
Technology, started work on a 3D visualization application, based on
Wingware's C++ Visualization Toolkit, called
Mayavi\footnote{https://web.archive.org/web/20030731122452/http://pythonology.org/success\&story=mayavi}.

SciPy initially contained a plotting module, but it did not have an
easy and familiar interface. \fixme{There were multiple plotting
  packages in SciPy, and Chaco, and wxPyPlot(s), IPython + GNUPlot,
  and ....?}
John Hunter, a gradute student
\fixme{check, since John around that time became postdoc, can check
  with John Milton?} at the University
of Chicago Medical School, liked the plotting functionality
available in MATLAB, but had problems accessing their laboratory's
license, which was governed by a hardware dongle.  In response, he
wrote a plotting library from scratch, and Matplotlib 0.1 was released
April 2003\footnote{https://mail.python.org/pipermail/python-list/2003-April/193167.html}.

In 2001, Fernando Pérez, a graduate student at the University of
Colorado, Boulder, created the IPython interactive shell, built into
IPP, an earlier attempt by Janko Hauser, and including code from
Nathan Gray's LazyPython.
% IPython hosted on SciPy:
% https://mail.python.org/pipermail/scipy-user/2003-March/001455.html

In 2000, STScI ported their entire IDL-based analysis pipeline, IRAF,
to Python.  And it was here that a small crisis developed around the
Python numerical array container.  Numeric, the original array package, was
suitable for small arrays, but not for the large images processed by
STScI.  With the Numeric maintainer's blessing, the decision was made
to write NumArray, a library that could handle data on a different
scale.  Unfortunately, NumArray proved inefficient for small arrays,
presenting the community with a rather unfortunate choice.  In 2005,
Travis Oliphant combined the best elements of Numeric and NumArray,
thereby solving the dilemma and paving the way for Python to become
a very significant player in the Data Science movement.\fixme{NumPy
  1.0 was released in October 2006}

@@

@@

CiSE 2007 special issue

@@

\cite{dubois2007guest}
\cite{oliphant2007python}
\cite{perez2007ipython}
\cite{hunter2007matplotlib}
\cite{vanderwalt2008scipy}
\cite{peterson2009f2py}
\cite{millman2011python}
\cite{perez2011python}
\cite{vanderwalt2011numpy}
\cite{behnel2011cython}
\cite{pedregosa2011scikit}
\cite{vanderwalt2014scikit}
\cite{meurer2017sympy}

http://conference.scipy.org/proceedings/scipy2008/paper\_0/full\_text.pdf
http://conference.scipy.org/proceedings/scipy2008/paper\_1/full\_text.pdf

The first SciPy workshop was held in September 2002 at CalTech—a
single track, two day event with 50 participants, many of them
developers of SciPy and surrounding libraries.  In the early
workshops, recurrent topics indicate the state of development, with
emphasis being placed on the underlying array package, plotting,
parallel processing, acceleration / wrapping, and user interfaces.  By
2004, a significant shift occurred towards application of SciPy on
scientific problems.  The event also started to draw in more keynote
speakers from outside the community, such as Guido van Rossum (creator
of Python, 2006), Ivan Krstić (One Laptop per Child, 2007), Alex
Martelli (Google, 2008), and Peter Norvig (Google Research, 2009).

%2002—https://web.archive.org/web/20021020215937/http://www.scipy.org:80/site_content/scipy02/scipy02_agenda.htm
%2003—https://web.archive.org/web/20031207151811/http://www.scipy.org:80/site_content/scipy03/
%2004—Jim Hugunin
https://web.archive.org/web/20040920080225/http://www.scipy.org:80/wikis/scipy04
%Travis Oliphant 2005—https://web.archive.org/web/20051103155042/http://www.scipy.org:80/wikis/scipy05/FrontPage
%Guido—2006 https://scipy.github.io/old-wiki/pages/SciPy2006.html
%Martelli—2008 https://scipy.github.io/old-wiki/pages/Developer_Zone/Conferences/SciPy2004.html
%Norvig—2009 https://scipy.github.io/old-wiki/pages/Developer_Zone/Conferences/SciPy2005.html

While the US SciPy conference kept growing (by ... it had multiple
tracks, student sponsorship, proceedings, and funding), satellite
events were being organized by volunteers elsewhere, such as EuroSciPy
(2008–) and SciPy India (2009–).

In 2008, the SciPy conference went from being a small gathering of
core developers, to a multi-location event, with increased funding, a
published proceedings, and scholarships for attending students (some
of whom are authors on this paper).  In 2010, the conference grew even
further, with an additional event added in India, and the US
conference—now with parallel tracks—moving to Austin, Texas.

>>>>

Initial work on numerical computing in Python was driven by graduate
students, but soon larger research labs became increasingly engaged.
For example, Paul Dubois at Lawrence Livermore National Laboratory took over the
maintenance of Numeric and funded the writing of its
manual\footnote{http://www.numpy.org/\_downloads/numeric-manual.pdf}, and
%in the Program for Climate Model Diagnosis and Intercomparison at
%Lawrence Livermore National Laboratory built their Climate Data
%Analysis Tools (``an open-source set of Python modules and tools for
%accessing, manipulating, and plotting scientific data in general and
%climate data sets in
%particular''\footnote{https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html})
%on Numeric, the newly established Python array object.  Paul later
%became the maintainer for Numeric, and helped fund the writing of the
%first manual\footnote{https://conference.scipy.org/scipy2010/slides/travis_oliphant_keynote.pdf}
in 1998 the Space Telescope Science Institute, in charge of science
operations of the Hubble Space Telescope, decided to replace their
custom scripting language with Python\footnote{https://conference.scipy.org/scipy2011/slides/greenfield\_keynote\_astronomy.pdf}.

Many of the first efforts were driven by graduate
students and post-doctoral researchers, but soon attracted the
attention of more senior researchers.
% Paul Dubois, CDAT
% https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html
For example,
% Perry Greenfield, Hubble
% https://conference.scipy.org/scipy2011/slides/greenfield_keynote_astronomy.pdf
%
% Note: plotting was important for them, so they eventually
% funded development on Matplotlib.

@@

With a rich programming environment and a numerical array object in
place, the time was ripe for the development of a full scientific
software stack.
In 2001, Eric Jones and Travis Vaught founded Enthought Scientific
Computing Solutions (now Enthought, Inc.) in Austin, Texas.  In an
effort to simplify the tool stack, they created the SciPy project,
centered around the SciPy library which would subsume all the
above-mentioned packages.
The new project quickly gained momentum, with a website and code
repository appearing in
February\footnote{https://web.archive.org/web/20010309040805/http://scipy.org:80/}, and
a mailing list announced in
June \footnote{https://mail.python.org/pipermail/scipy-dev/2001-June/000000.html}. By
August, a first release was announced\footnote{https://mail.python.org/pipermail/python-list/2001-August/106419.html}, an excerpt of which is shown in
Fig.~\ref{fig:announce-0.1}.
In September, the first documentation was
published\footnote{https://web.archive.org/web/20021013204556/http://www.scipy.org:80/scipy/site\_content/site\_news/docs\_released1}.
The first SciPy
workshop\footnote{https://mail.python.org/pipermail/numpy-discussion/2002-June/001511.html}
was held in September 2002 at CalTech---a single track, two day event with 50
participants, many of them developers of SciPy and surrounding libraries.

%and by the following September the inaugural SciPy workshop was held
%at Caltech\footnote{https://mail.python.org/pipermail/numpy-discussion/2002-June/001511.html}.

@@

At this point, scientific Python started attracting more serious attention;
code that started out as side projects by graduate students has grown into
essential infrastructure at national laboratories and research institutes.
For example, starting in 2000, the Space Telescope Science Institute (STScI)
ported their entire IDL-based analysis pipeline, IRAF, to Python.
STScI was founded to guide the Hubble Space Telescope including developing
advanced data, archives, and analysis tools for the most famous observatory
in history.

As SciPy, the algorithms library of the ecosystem, began attracting the attention
of large research organizations,
the next generation of graduate students and postdocs were already exploring
other aspects of the computational environment.
In 2000, Prabhu Ramachandran\fixme{also note that he, and the others
  mentioned below, are also SciPy contributors}, a PhD student at the Indian Institute of
Technology, started work on a 3D visualization application, based on
Wingware's C++ Visualization Toolkit, called
Mayavi\footnote{https://web.archive.org/web/20030731122452/http://pythonology.org/success\&story=mayavi}.
In 2001, Fernando Pérez, a graduate student at the University of
Colorado, Boulder, created the IPython interactive shell, built into
IPP, an earlier attempt by Janko Hauser, and including code from
Nathan Gray's LazyPython.
% IPython hosted on SciPy:
% https://mail.python.org/pipermail/scipy-user/2003-March/001455.html
John Hunter, a gradute student
\fixme{check, since John around that time became postdoc, can check
  with John Milton?} at the University
of Chicago Medical School, liked the plotting functionality
available in MATLAB, but had problems accessing their laboratory's
license, which was governed by a hardware dongle.  In response, he
wrote a plotting library from scratch, and Matplotlib 0.1 was released
April 2003\footnote{https://mail.python.org/pipermail/python-list/2003-April/193167.html}.

In the early
workshops, recurrent topics indicate the state of development, with
emphasis being placed on the underlying array package, plotting,
parallel processing, acceleration / wrapping, and user interfaces.  By
2004, a significant shift occurred towards application of SciPy on
scientific problems.  The event also started to draw in more keynote
speakers from outside the community, such as Guido van Rossum (creator
of Python, 2006), Ivan Krstić (One Laptop per Child, 2007), Alex
Martelli (Google, 2008), and Peter Norvig (Google Research, 2009).

%2002—https://web.archive.org/web/20021020215937/http://www.scipy.org:80/site_content/scipy02/scipy02_agenda.htm
%2003—https://web.archive.org/web/20031207151811/http://www.scipy.org:80/site_content/scipy03/
%2004—Jim Hugunin
%https://web.archive.org/web/20040920080225/http://www.scipy.org:80/wikis/scipy04
%Travis Oliphant 2005—https://web.archive.org/web/20051103155042/http://www.scipy.org:80/wikis/scipy05/FrontPage
%Guido—2006 https://scipy.github.io/old-wiki/pages/SciPy2006.html
%Martelli—2008 https://scipy.github.io/old-wiki/pages/Developer_Zone/Conferences/SciPy2004.html
%Norvig—2009 https://scipy.github.io/old-wiki/pages/Developer_Zone/Conferences/SciPy2005.html

As STScI continued to use Python for an increasingly large portion
of the Hubble Space Telescope data analysis pipeline, they encountered
problems with the Python numerical array container.
Numeric, the original array package, was
suitable for small arrays, but not for the large images processed by
STScI.  With the Numeric maintainer's blessing, the decision was made
to write NumArray, a library that could handle data on a different
scale.  Unfortunately, NumArray proved inefficient for small arrays,
presenting the community with a rather unfortunate choice.  In 2005,
Travis Oliphant combined the best elements of Numeric and NumArray,
thereby solving the dilemma and paving the way for Python to become
a very significant player in the Data Science movement.\fixme{NumPy
  1.0 was released in October 2006}

@@

CiSE 2007 special issue (steering code)
\cite{dubois2007guest}
\cite{oliphant2007python}
\cite{perez2007ipython}
\cite{hunter2007matplotlib}
research applications ...

@@

%http://conference.scipy.org/proceedings/scipy2008/paper\_0/full\_text.pdf
%http://conference.scipy.org/proceedings/scipy2008/paper\_1/full\_text.pdf

In 2008, the SciPy conference went from being a small gathering of
core developers, to a multi-location event, with increased funding, a
published proceedings, and scholarships for attending students (some
of whom are authors on this paper).  In 2010, the conference grew even
further, with an additional event added in India, and the US
conference—now with parallel tracks—moving to Austin, Texas.
While the US SciPy conference kept growing (by ... it had multiple
tracks, student sponsorship, proceedings, and funding), satellite
events were being organized by volunteers elsewhere, such as EuroSciPy
(2008–) and SciPy India (2009–).

@@

\cite{vanderwalt2008scipy}
\cite{peterson2009f2py}
\cite{millman2011python}
\cite{perez2011python}
\cite{vanderwalt2011numpy}
\cite{behnel2011cython}
\cite{pedregosa2011scikit}
\cite{vanderwalt2014scikit}
\cite{meurer2017sympy}
